Spectroscopic confirmation of very-wide stellar binaries and large-separation comoving pairs from \textsl{Gaia DR1}

### 2017-10-09 paper/comoving-followup.tex
Addressing @smoh comments

Of the 323 observed star pairs, 12 have at least one member with an unreliable
(flagged) radial velocity measurement from the spectroscopic reduction pipeline.
For the remaining 311 pairs, we construct a statistical model to determine the
probability that a given candidate comoving pair is truly comoving.
For this inference, the data are the proper motions in the ICRS reference frame,
$\bs{\mu}$, and parallax, $\varpi$, (from the \tgas\ catalog) and barycentric
radial velocity, $v_r$, for each star in a candidate comoving pair; we assume
that the sky position for each star $(\alpha, \delta)$ is known perfectly.
For brevity in the expressions below, we define $\bs{\eta} = (\mu_{\alpha^*},
\mu_\delta, v_r)$, where, e.g., $\bs{\eta}_1$ refers to the proper motion and
radial velocity data for star 1 in the pair.
To simplify some expressions, we additionally ignore any correlations between
the observed parallax and the proper motion components (but keep the covariance
structure between the two proper motion components).

Our model is expressed in terms of the magnitude of the true (unobserved) 3D
velocity difference for each pair, $\psi = \left| \bs{v}_1 - \bs{v}_2 \right|$,
and the (global) fraction of pairs that are comoving, $f$.
For the distribution of true 3D velocity difference magnitudes, $p(\psi \given
f)$, we use a mixture model with two components
\begin{eqnarray}
    p(\psi \given f) &=& f\,p_{\rm co}(\psi) + (1-f)\,p_{\rm field}(\psi)
\end{eqnarray}
where $p_{\rm co}$, the probability distribution function (pdf) of true comoving
pairs, and $p_{\rm field}$, the pdf of randomly paired field stars.
Using a mixture model and inferring the mixture weight is a common trick to
avoid explicitly having to sample over the per-pair probabilities: this
expression is equivalent to the marginal likelihood under a model in which every
candidate comoving pair has an additional latent parameter (a binary flag,
$q_n$) that specifies which mixture component it belongs to
(\citealt{Hogg:2010,Foreman-Mackey:2014}).
We later use this fact to compute the probability that each pair is truly
comoving, or, more precisely, belongs to the mixture component $p_{\rm co}$.

The form for both mixture components ($p_{\rm co}$ and $p_{\rm field}$) is taken
to be the probability of the magnitude of the 3D separation of two vectors under
the assumption that the two vectors were drawn independently from an isotropic,
3D Gaussian with dispersion $\sigma$.
That is,
\begin{eqnarray}
    f(\psi \,;\, \sigma) &=& \left[2\sqrt {\pi}\, \sigma^3 \right]^{-1} \,
        \psi^2 \, \ \exp\left[-\frac {\psi^2} {4\, \sigma^2} \right]
        \label{eq:vel-diff-dist}\\
    \bs{v_1}, \bs{v_2} &\sim& \mathcal{N}(\bs{0}, \sigma^2\,\mat{I}_3) \\
    p_{\rm co}(\psi) &=& f(\psi \,;\, \sigma = 1~\kms)\\
    p_{\rm field}(\psi) &=& f(\psi \,;\, \sigma = \sigma_{\rm field})
\end{eqnarray}
where $\sigma_{\rm field}$ is defined below, $\mathcal{N}(\bs{m},
\mat{\Sigma})$ represents a normal distribution with mean vector $\bs{m}$ and
covariance matrix $\mat{\Sigma}$, and $\mat{I}_3$ is the 3 by 3 identity matrix.

Under this model, the joint probability of the velocity data for a given pair
$\bs{\eta}_1$ and $\bs{\eta}_2$, and the latent parameters $\bs{v}_1, \bs{v}_2,
\psi$, conditional on the unobserved true distances $d_1$ and $d_2$ (we handle
these separately below), is
\begin{equation}
p(\bs{\eta}_1, \bs{\eta}_2, \bs{v}_1, \bs{v}_2, \psi \given d_1, d_2, f) =
    p(\bs{\eta}_1 \given d_1, \bs{v}_1) \,
    p(\bs{\eta}_2, v_{r,2} \given d_2, \bs{v}_2) \,
    p(\bs{v}_1, \bs{v}_2 \given \psi) \, p(\psi \given f)\label{eq:full-ll}
\end{equation}
where
\begin{eqnarray}
    p(\bs{v}_1, \bs{v}_2 \given \psi) &=& \delta(|\bs{v}_1-\bs{v}_2| - \psi) \quad .
\end{eqnarray}

Because we are only interested in inferring the fraction of stars consistent
with being comoving, we marginalize this expression over the true 3D velocities
and velocity difference, obtaining the velocity-marginalized, per-pair
likelihood
\begin{equation}
    p(\bs{\eta}_1, \bs{\eta}_2 \given d_1, d_2, f) =
    \int \dd^3\bs{v}_1 \, \dd^3\bs{v}_2 \, \dd \psi \,
    p(\bs{\eta}_1, \bs{\eta}_2, \bs{v}_1, \bs{v}_2, \psi \given d_1, d_2, f)
    \quad .
\end{equation}
This expression is still conditioned on the true distances, $d_1$ and $d_2$, but
we observe the parallaxes $\varpi_1$ and $\varpi_2$.
To remove the final per-pair latent parameters in this model, we therefore also
have to marginalize the likelihood over these distance parameters to obtain the
velocity-and-distance marginalized likelihood
\begin{equation}
    p(\bs{\eta}_1, \bs{\eta}_2, \varpi_1, \varpi_2 \given f) =
        \int \dd d_1 \, \dd d_2 \,
        p(\varpi_1 \given d_1) \, p(\varpi_2 \given d_2) \,
        p(\bs{\eta}_1, \bs{\eta}_2 \given d_1, d_2, f) \,
        p(d_1) \, p(d_2)
\end{equation}
where we assume a uniform space-density prior for the distances
(\citealt{Bailer-Jones:2015aa}) with a maximum distance of 1~kpc.

The posterior probability over the fraction of true comoving star pairs, $f$, is
then given by
\begin{equation}
    p(f \given \{\bs{\eta}_1, \bs{\eta}_2, \varpi_1, \varpi_2\}_{N}) \propto
        p(f) \, \prod_n^N
        p(\bs{\eta}_{1,n}, \bs{\eta}_{2,n}, \varpi_{1,n}, \varpi_{2,n} \given f)
\end{equation}
where $N$ is the number of candidate comoving star pairs and the index $n$
specifies the data for the $n$th pair.

We approximate all of the marginalization integrals using numerical procedures.
Given distances $d_1$ and $d_2$, we can approximate the velocity marginalization
using Monte Carlo integration:
\begin{equation}
    p(\bs{\eta}_1, \bs{\eta}_2 \given d_1, d_2, f) \approx
    \frac{1}{K} \, \sum^K_k \,
    p(\psi^{(k)} \given f) \quad .
\end{equation}
In this expression, the $k$-th sample $\psi^{(k)}$ is defined as
\begin{eqnarray}
    \psi^{(k)} &=& |\bs{v}_1^{(k)} - \bs{v}_2^{(k)}| \\
    \bs{v}_1^{(k)} &=& \mat{M}(\alpha_1, \delta_1) \cdot
        \left(\begin{array}{c}
            d_1\,\mu_{\alpha^*, 1}^{(k)} \\
            d_1\,\mu_{\delta, 1}^{(k)} \\
            v_{r, 1}^{(k)}
        \end{array}\right)\\
    \left(\begin{array}{c}
        \mu_{\alpha^*, 1}^{(k)} \\
        \mu_{\delta, 1}^{(k)} \\
        v_{r, 1}^{(k)}
    \end{array}\right) &\sim& \mathcal{N}(\bs{\eta}_1, \mat{C}_1)
\end{eqnarray}
where $\mat{C}_1$ is the covariance matrix of the \gaia\ data provided by the
\tgas\ catalog with an additional (uncorrelated) diagonal term added for the
uncertainty in the radial velocity measurement, and the rotation matrix
$\mat{M}(\alpha, \delta)$ transforms from the local, tangent-space coordinates
at $(\alpha, \delta)$ to the ICRS (see, e.g., Section~3 of \citealt{Oh:2017}).
We perform the outer integral over the two distances, $d_1$ and $d_2$, using
Simpson's rule on a 2D grid of distances that, in each dimension, spans the
domain $\left[ (\varpi + 3\sigma_{\varpi})^{-1}, (\varpi -
3\sigma_{\varpi})^{-1} \right]$, where $\sigma_{\varpi}$ is the parallax
uncertainty reported in the \tgas\ catalog for each star.
From convergence tests, we have found that it is sufficient to use 1024 samples
in the Monte Carlo integral and 5 by 5 distance grid points.

We use an isotropic model for the field velocity distribution with dispersion
set by taking the trace of the local velocity dispersion tensor for young
($<2~{\rm Gyr}$) thin disk stars from \citealt{Nordstrom:2004}, $\mat{V}_{\rm
N04}$, $\sigma_{\rm field} =
\sqrt{\frac{1}{3}\Tr{\left(\mat{V}_{\rm N04}\right)}} \approx
18~\kms$.\footnote{Even though the dispersion in this isotropic model for the
field stars is only $18~\kms$, most randomly paired stars will tend to have 3D
velocity differences $>20~\kms$. The 15th and 85th percentiles of the
distribution of velocity differences (\eqname~\ref{eq:vel-diff-dist}) are
$\approx 23~\kms$ and $\approx 59~\kms$, respectively.}
This is a conservative choice: the velocity dispersion of thin disk stars
increases steadily with age (e.g., \citealt{Aumer:2016}), so assuming that all
of our pairs are young should lead to under-predictions of the probability that
a given pair is truly comoving.
A more robust model that simultaneously models the ages and age distribution of
the comoving pairs would lead to more precise inferences of these probabilities,
but is outside of the scope of this work.

>>>>

Of the 323 observed star pairs, 12 have at least one member with an unreliable
(flagged) radial velocity measurement from the spectroscopic reduction pipeline.
For the remaining 311 pairs, we construct a statistical model to determine the
probability that a given candidate comoving pair is comoving in 3D velocity
given the radial velocity and uncertainty measured in this work combined with
\tgas\ astrometric data.
For this inference, the data are the proper motions in the ICRS reference frame,
$\bs{\mu} = (\mu_{\alpha^*}, \mu_\delta)$,\footnote{$\mu_{\alpha^*} = \mu_\alpha
\, \cos\delta$} and parallax, $\varpi$, (from the
\tgas\ catalog) and barycentric radial velocity, $v_r$, for each star in a
candidate comoving pair; we assume that the sky position for each star $(\alpha,
\delta)$ is known perfectly.
For brevity in the expressions below, we define $\bs{\eta} = (\mu_{\alpha^*},
\mu_\delta, v_r)$, where, e.g., $\bs{\eta}_1$ refers to the proper motion and
radial velocity data for star 1 in the pair.
To simplify some expressions, we additionally ignore any correlations between
the observed parallax and the proper motion components (but keep the covariance
between the two proper motion components).

Our model is expressed in terms of the magnitude of the true (latent) 3D
velocity difference for each pair, $\psi = \left| \bs{v}_1 - \bs{v}_2 \right|$,
and the (global) fraction of pairs that are comoving, $f$.
For the distribution of true 3D velocity difference magnitudes, $p(\psi \given
f)$, we use a mixture model with two components
\begin{eqnarray}
    p(\psi \given f) &=& f\,p_{\rm co}(\psi) + (1-f)\,p_{\rm field}(\psi)
\end{eqnarray}
where $p_{\rm co}$, the probability distribution function (pdf) of true comoving
pairs, and $p_{\rm field}$, the pdf of randomly paired field stars.
Using a mixture model and inferring the mixture weight is a trick to avoid
explicitly having to sample over the probabilities for each individual object
(candidate comoving pair): this expression is equivalent to the marginal
likelihood under a model in which every candidate comoving pair has an
additional latent parameter (a binary flag, $q_n$) that specifies which mixture
component it belongs to (\citealt{Hogg:2010,Foreman-Mackey:2014}).
We later use this fact to compute the probability that each pair is truly
comoving, or, more precisely, belongs to the mixture component $p_{\rm co}$.

The form for both mixture components ($p_{\rm co}$ and $p_{\rm field}$) is taken
to be the probability distribution of the magnitude of the 3D difference between
two vectors under the assumption that the two vectors were drawn independently
from an isotropic, 3D Gaussian with dispersion $\sigma$.
For the three-dimensional velocity vectors we consider here, the distribution is
Maxwellian in form.
That is,
\begin{eqnarray}
    f(\psi \,;\, \sigma) &=& \left[2\sqrt {\pi}\, \sigma^3 \right]^{-1} \,
        \psi^2 \, \ \exp\left[-\frac {\psi^2} {4\, \sigma^2} \right]
        \label{eq:vel-diff-dist}\\
    \bs{v_1}, \bs{v_2} &\sim& \mathcal{N}(\bs{v} \given \bs{0}, \sigma^2\,\mat{I}_3) \\
    p_{\rm co}(\psi) &=& \mathcal{N}(\psi \given 0, \sigma^2_{\rm co})\\
    p_{\rm field}(\psi) &=& f(\psi \,;\, \sigma_{\rm field})
\end{eqnarray}
where $\sigma_{\rm co}$ and $\sigma_{\rm field}$ are defined below,
$\mathcal{N}(\bs{v} \given \bs{m}, \mat{\Sigma})$ represents a normal
distribution over $\bs{v}$ with mean vector $\bs{m}$ and covariance matrix
$\mat{\Sigma}$, $\mathcal{N}(x \given m, \sigma^2)\,[a,b]$ is a truncated normal
distribution with mean $m$ and variance $\sigma^2$ defined on the domain
$[a,b]$, and $\mat{I}_3$ is the 3 by 3 identity matrix.

Under this model, the joint probability of the velocity data for a given pair
$\bs{\eta}_1$ and $\bs{\eta}_2$, and the latent parameters $\bs{v}_1, \bs{v}_2,
\psi$, conditional on the unobserved true distances $d_1$ and $d_2$ (we handle
these separately below), is
\begin{equation}
p(\bs{\eta}_1, \bs{\eta}_2, \bs{v}_1, \bs{v}_2, \psi \given d_1, d_2, f) =
    p(\bs{\eta}_1 \given d_1, \bs{v}_1) \,
    p(\bs{\eta}_2 \given d_2, \bs{v}_2) \,
    p(\bs{v}_1, \bs{v}_2 \given \psi) \, p(\psi \given f)\label{eq:full-ll}
\end{equation}
where
\begin{eqnarray}
    p(\bs{v}_1, \bs{v}_2 \given \psi) &=& \delta(|\bs{v}_1-\bs{v}_2| - \psi) \quad .
\end{eqnarray}

Because we are only interested in inferring the fraction of stars consistent
with being comoving, $f$, we marginalize this expression over the true 3D
velocities and velocity difference, obtaining the velocity-marginalized,
per-pair likelihood
\begin{equation}
    p(\bs{\eta}_1, \bs{\eta}_2 \given d_1, d_2, f) =
    \int \dd^3\bs{v}_1 \, \dd^3\bs{v}_2 \, \dd \psi \,
    p(\bs{\eta}_1, \bs{\eta}_2, \bs{v}_1, \bs{v}_2, \psi \given d_1, d_2, f)
    \quad . \label{eq:vel-marg}
\end{equation}
This expression is still conditioned on the true distances, $d_1$ and $d_2$, but
we observe the parallaxes $\varpi_1$ and $\varpi_2$.
To remove the final per-pair latent parameters in this model, we therefore also
have to marginalize the likelihood over these distance parameters to obtain the
velocity-and-distance marginalized likelihood
\begin{equation}
    p(\bs{\eta}_1, \bs{\eta}_2, \varpi_1, \varpi_2 \given f) =
        \int \dd d_1 \, \dd d_2 \,
        p(\bs{\eta}_1, \bs{\eta}_2 \given d_1, d_2, f) \,
        p(\varpi_1 \given d_1) \, p(\varpi_2 \given d_2) \,
        p(d_1) \, p(d_2)
\end{equation}
where we assume a uniform space-density prior for the distances
(\citealt{Bailer-Jones:2015aa}) with a maximum distance of 1~kpc.

The posterior probability over the fraction of true comoving star pairs, $f$, is
then given by
\begin{equation}
    p(f \given \{\bs{\eta}_{n1}, \bs{\eta}_{n2}, \varpi_{n1}, \varpi_{n2}\}_N)
        \propto p(f) \, \prod_n^N
        p(\bs{\eta}_{n1}, \bs{\eta}_{n2}, \varpi_{n1}, \varpi_{n2} \given f)
\end{equation}
where $(\bs{\eta}_{n1}, \varpi_{n1})$ and $(\bs{\eta}_{n2}, \varpi_{n2})$ refer
to the first and second star in the $n$-th comoving pair, respectively, and  $N$
is the total number of candidate comoving star pairs.

We approximate all of the marginalization integrals using numerical procedures.
Given distances $d_1$ and $d_2$, we estimate the velocity marginalization
(\eqname~\ref{eq:vel-marg}) using Monte Carlo integration:
\begin{equation}
    p(\bs{\eta}_1, \bs{\eta}_2 \given d_1, d_2, f) \approx
    \frac{1}{K} \, \sum^K_k \,
    p(\psi^{(k)} \given f) \quad .
\end{equation}
In this expression, the $k$-th Monte Carlo sample $\psi^{(k)}$ is defined as
\begin{eqnarray}
    \psi^{(k)} &=& |\bs{v}_1^{(k)} - \bs{v}_2^{(k)}| \\
    \bs{v}_1^{(k)} &=& \mat{M}(\alpha_1, \delta_1) \cdot
        \left(\begin{array}{c}
            d_1\,\mu_{\alpha^*, 1}^{(k)} \\
            d_1\,\mu_{\delta, 1}^{(k)} \\
            v_{r, 1}^{(k)}
        \end{array}\right)\label{eq:v1-sample} \\
    \left(\begin{array}{c}
        \mu_{\alpha^*, 1}^{(k)} \\
        \mu_{\delta, 1}^{(k)} \\
        v_{r, 1}^{(k)}
    \end{array}\right) &\sim& \mathcal{N}(\bs{\eta}_1, \mat{C}_1) \label{eq:mu1-sample}
\end{eqnarray}
where $\mat{C}_1$ is the covariance matrix of the \gaia\ data provided by the
\tgas\ catalog with an additional (uncorrelated) diagonal term added for the
uncertainty in the radial velocity measurement, and the rotation matrix
$\mat{M}(\alpha, \delta)$ transforms from the local, tangent-space coordinates
at $(\alpha, \delta)$ to the ICRS (see, e.g., Section~3 of \citealt{Oh:2017}).
Expressions for star 2 are equivalent with index 1 replaced with 2 in \eqname
s~\ref{eq:v1-sample}--\ref{eq:mu1-sample}.
We perform the outer integral over the two distances, $d_1$ and $d_2$, using
Simpson's rule on a 2D grid of distances that, in each dimension, spans the
domain $\left[ (\varpi + 3\sigma_{\varpi})^{-1}, (\varpi -
3\sigma_{\varpi})^{-1} \right]$,\footnote{All of our stars have high enough
parallax signal-to-noise ratios that integration over this grid is equivalent to
integrating over much larger domains. We tested this by increasing the grid to
$\left[ (\varpi + 10\sigma_{\varpi})^{-1}, (\varpi - 10\sigma_{\varpi})^{-1}
\right]$ and comparing the integral estimates.} where $\sigma_{\varpi}$ is the
parallax uncertainty reported in the \tgas\ catalog for each star.
From convergence tests, we have found that it is sufficient to use 1024 samples
in the Monte Carlo integral and 5 by 5 distance grid points.

We use an isotropic model for the field velocity distribution with dispersion
set by taking the trace of the local velocity dispersion tensor for young
($<2~{\rm Gyr}$) thin disk stars from \citealt{Nordstrom:2004}, $\mat{V}_{\rm
N04}$, $\sigma_{\rm field} =
\sqrt{\frac{1}{3}\Tr{\left(\mat{V}_{\rm N04}\right)}} \approx
18~\kms$.\footnote{Even though the dispersion in this isotropic model for the
field stars is only $18~\kms$, most randomly paired stars will tend to have 3D
velocity differences $>20~\kms$. The 15th and 85th percentiles of the
distribution of velocity differences (\eqname~\ref{eq:vel-diff-dist}) are
$\approx 23~\kms$ and $\approx 59~\kms$, respectively.
This is related to the fact that, for a three-dimensional Gaussian, most of the
probability \emph{mass} is not at $\bs{0}$, but is rather concentrated in an
annular region at finite distance from the origin.}
This is a conservative choice: the velocity dispersion of thin disk stars
increases steadily with age (e.g., \citealt{Aumer:2016}), so assuming that all
of our pairs are young should lead to under-predictions of the probability that
a given pair is truly comoving.
A more robust model that simultaneously models the ages and age distribution of
the comoving pairs would lead to more precise inferences of these probabilities,
but is outside of the scope of this work.

At the smallest tangential separations of our candidate pairs, ($s_{\rm tan}
\approx 10^{-3}~\pc$), if the two stars are bound, the expected velocity
difference is a few $\kms$ (depending on mass).
We therefore set $\sigma_{\rm co} = 2~\kms$
